Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2023-01-10 response_to_reviewers_R1.Rmd

> We have fixed the wording of this sentence. Instead of "extend", it now says "consider a similar approach".

- Also in Section 2.1.2, it might be worthwhile to briefly define absorbing states, as I for
one didn’t know the definition before looking it up (and as it turns out the concept is
quite simple)

> We have added a sentence defining what an absorbing state is: "An absorbing state is a link in a Markov chain from which there are no possible transitions (i.e., it is the end of the chain)."

- In Section 2.1.3, you make the comment “they estimate that a better strategy would
be worth about 0.4 wins per season on average, a substantial amount in a season with
only 16 games.” This makes sense, but previously in Section 2.1.2 you mentioned
a potential difference of 3.8 wins per season, yet didn’t comment on the significant
magnitude of that effect. I’m not advocating for a particular change here, but the
authors could consider adding a small comment about the size of the effect Pelechrinis
et. al. estimated.

> We have moved the comment up and connected the second comment to the first.

- At the end of Section 2.1.3, you write “Lopez (2020) also concludes that coaches are
too conservative on 4th down decision making, but the effect may be overstated due
to the way in which to go yardage has been recorded in the past.” It’s unclear to me
whether the overstated effect is the one Lopez estimates in his paper, or if you are
referring to the estimated effects past studies which used less granular data.

> We have clarified this language. It now reads: "Lopez (2020) also concludes that coaches are too conservative on 4th down decision making, but notes further that past estimates of the magnitude of how conservative coaches are on 4th down may be overstated due to the way in which to go yardage was recorded only approximately in the past."

- In Section 2.3, you write “NHL coaches do not always employ the optimal strategies”,
but it isn’t clear what those strategies are in reference to the previous discussion. Is
pulling the goalie optimal?

### 2023-01-10 sports_analytics.Rmd
close #49

@lopez2020bigger presents an introduction to NFL tracking data, and examines 4th down behavior as an example of the type of problem that can be more thoroughly studied with the increase in granularity of the tracking data over traditional NFL data.
In the past, when looking at down and distance data to study whether NFL coaches are making good decisions about whether to "go for it" or punt on 4th down, the distance data is only a rounded approximation of the true distance "to go" (i.e. 1 yard, 1 foot, and 1 inch will all be recorded as 4th and 1. 
In fact, anything up to 2 yard will recorded as 4th and 1 [@lopezDM]. However, a coach on the field during a game will be able to clearly see the difference between 1 inch and 1 yard, and this information will factor into their decision making. 
With tracking data, the "to go" distance can be much more accurately assessed and therefore evaluation of 4th down coaching decisions can now account for this "extra" information that is available to a coach on the field of play, but not recorded in traditional NFL data. 
Many past analyses of the decision to go for it or not on 4th down conclude that coaches in the NFL are too conservative in their decision making. 
@lopez2020bigger also concludes that coaches are too conservative on 4th down decision making, but further notes that past estimates of the magnitude of how conservative coaches are on 4th down may be overstated due to the way in which to go yardage was recorded only approximately in the past.

@@

## Example of how to use betting market data

To see how betting market data can be used to estimate team strengths, consider the betting odds posted on [FanDuel Sportsbook for the 2023 NBA Champion](https://sportsbook.fanduel.com/navigation/nba?tab=nba-finals), posted on January 9, 2023 and shown in Figure \ref{tab:odds}. 
This is a futures market, because the actual NBA champion will not be determined until June 2023. 
The Boston Celtics are the favorite to win, with posted odds of $+390$, meaning that a \$100 bet on the Celtics to win the championship will pay out \$390 if the Celtics win it all. 
Conversely, six teams share the lowest odds at $+50000$.

```{r odds}
library(tidyverse)
odds <- read_csv("odds.csv")
odds |>
  mutate(
    prob = 100 / (100 + line),
    `Prob. Normalized` = prob / sum(prob)
  ) |>
  rename(Team = team, Line = line, `Prob.` = prob) |>
  mutate(Rank = row_number(), .before = Team) |>
  janitor::adorn_totals() |>
  slice(c(1:6, 25:31)) |>
  knitr::kable(
    booktabs = TRUE, 
    linesep = c('', '', '', '', '', '\\addlinespace'),
    digits = 3, 
    caption = "2023 NBA Championship line odds for the top 6 and bottom 6 teams. Retrieved from FanDuel Sportsbook on January 9, 2023. ")
```

These lines ($\ell_i$) can be converted into an implied probability ($p_i$) using the formula:
$$
  p_i = \frac{100}{100 + \ell_i} \,.
$$
The sum of those probabilities is greater than one---this is why the sportsbook makes money regardless of who wins the championship. 
However, the implied probabilities can be normalized by dividing by their sum to recover true probabilities of each team winning the championship. 
Many different researchers have shown that these normalized implied probabilities are accurate, unbiased, and efficient estimates of the true unknowable probabilities (see @Lopez2018 for discussion and an extensive list of references).

>>>>

@lopez2020bigger presents an introduction to NFL tracking data, and examines 4th down behavior as an example of the type of problem that can be more thoroughly studied with the increase in granularity of the tracking data over traditional NFL data.
In the past, when looking at down and distance data to study whether NFL coaches are making good decisions about whether to "go for it" or punt on 4th down, the distance data is only a rounded approximation of the true distance "to go" (i.e. 1 yard, 1 foot, and 1 inch will all be recorded as 4th and 1. 
In fact, anything up to 2 yard will recorded as 4th and 1 [@lopezDM]. However, a coach on the field during a game will be able to clearly see the difference between 1 inch and 1 yard, and this information will factor into their decision making. 
With tracking data, the "to go" distance can be much more accurately assessed and therefore evaluation of 4th down coaching decisions can now account for this "extra" information that is available to a coach on the field of play, but not recorded in traditional NFL data. 
Many past analyses of the decision to go for it or not on 4th down conclude that coaches in the NFL are too conservative in their decision making. 
@lopez2020bigger also concludes that coaches are too conservative on 4th down decision making, but notes further that past estimates of the magnitude of how conservative coaches are on 4th down may be overstated due to the way in which to go yardage was recorded only approximately in the past.

@@

## Example: Win probabilities from betting market data

To see how betting market data can be used to estimate team strengths, consider the betting lines posted on [FanDuel Sportsbook for the 2023 NBA Champion](https://sportsbook.fanduel.com/navigation/nba?tab=nba-finals) on January 9, 2023 and shown in Table \ref{tab:odds}. 
This is a futures market, because the actual NBA champion will not be determined until June 2023. 
The Boston Celtics are the favorite to win, with a moneyline of $+390$, meaning that a \$100 bet on the Celtics to win the championship will pay back the original bet and an additional \$390 if the Celtics win it all. 
This style of odds are sometimes called [*American odds*](https://en.wikipedia.org/wiki/Fixed-odds_betting#Moneyline_odds). 
The corresponding fractional odds have the Celtics at 4.9:1 to win the championship. 
Conversely, six teams share the lowest odds at $+50000$.

```{r odds}
library(tidyverse)
odds <- read_csv("odds.csv")
odds |>
  mutate(
    Odds = (line / 100) + 1,
    prob = 100 / (100 + line),
    `Prob. Normalized` = prob / sum(prob)
  ) |>
  rename(Team = team, Line = line, `Prob.` = prob) |>
  mutate(Rank = row_number(), .before = Team) |>
  janitor::adorn_totals() |>
  slice(c(1:6, 25:31)) |>
  knitr::kable(
    booktabs = TRUE, 
    linesep = c('', '', '', '', '', '\\addlinespace'),
    digits = 3, 
    caption = "2023 NBA Championship odds for the top 6 and bottom 6 teams. Retrieved from FanDuel Sportsbook on January 9, 2023. ")
```

These moneylines ($\ell_i$) can be converted into an implied probability ($p_i$) using the formula:
$$
  p_i = \frac{100}{100 + \ell_i} \,.
$$
The sum of those probabilities is greater than one---this is why the sportsbook makes money regardless of who wins the championship. 
However, the implied probabilities can be normalized by dividing by their sum to recover true probabilities of each team winning the championship. 
Many different researchers have shown that these normalized implied probabilities are accurate, unbiased, and efficient estimates of the true unknowable probabilities (see @Lopez2018 for discussion and an extensive list of references).
